\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-03-21 Articles/2020-QuantumLogic/Source/quantumlogic.tex
Another pass at the math section

\section{Math}
\label{Math}
In this section, we want to show how Hilbert spaces already are equipped with a mathematical structure that follows classical logic, namely the Borel algebra. We will see that the algebra typically associated with quantum logic is a subset of this structure. We will see that this classical structure contains other statements that are experimentally significant, like statements about the expectation values. Finally, we will show that the algebra of subspaces of classical distributions is also non-distributive in the same way that quantum logic is.

Let us start with a brief summary of the basic mathematical structure of quantum logic. Each proposition for a Hilbert space $H$ correspond to a closed subspace. This forms a complete lattice under the following definitions: the ordering $p \leq q$ of the lattice (i.e. the implication $p \to q$) represents subspace inclusion, the conjunction $p \wedge q$ is the biggest subspaces contained by both, the disjunction $p \vee q$ is the smallest subspace that contains both, the negation $\neg p$ is the orthogonal subspace.

To understand how the previous structure fails to a be Boolean algebra, we note that by Stone's representation theorem any Boolean algebra can be expressed as an algebra of sets using the standard set operations (i.e. inclusion, intersection, union and complement). If we understand each subspace as a set of vectors, the disjunction $p \vee q$ is not the mere set union of the subspaces, but the span. This is why the above structure, even if it is a complemented $0-1$ lattice, it fails to be a Boolean alebgra.

However, a Hilbert space $H$ is a complete normed space. The norm induces a topology, which in turns induces a Borel-algebra. The Borel-algebra is a family of sets closed under countable intersection, union and negation. It therefore forms a Boolean algebra. Therefore propositions of the type ``the state is in set $A$'' for any Borel set $A$ follow the rules of classical logic.

Note that the lattice of quantum logic is formed by closed subspaces, which are closed sets and therefore Borel sets. Therefore all statements expressible in quantum logic (e.g. ``we measured/prepared this quantity with the following value'') are also expressible with the Borel algebra. The converse, however, is not true: not all Borel sets correspond to propositions in quantum logic.

To show that some of these statements are indeed interesting, let us consider the following result:

\begin{prop}
	Let $H$ be an Hilbert space and $\langle \, , \, \rangle$ be its inner product. Let $X$ be a self-adjoint bounded linear operator. Let $F_X : H \to \mathbb{R}$ the map defined by $F_X(\psi) = \langle \psi , X \psi \rangle$. Let $U \subseteq \mathbb{R}$ be a Borel set. Then $F_X^{-1}(U)$ is a Borel set.
\end{prop}

A proof would work along these lines. A linear operator between two normed spaces is continuous if and only if it is bounded, therefore our operator $X$ is continuous. Since the inner product is also continuous, $F_X$ is continuous. Since $F_X$ is continuous, the reverse image of an open set is an open set. Since a Borel set $U$ can be constructed from open sets using negation and countable union, the reverse image $F_X^{-1}(U)$ can also be constructed from open sets using negation and countable union, which means it is a Borel set.

@@

Borel sets allow us to express statements such as ``the expected position $\langle x \rangle$ of the particle is exactly $x_0$''. This is NOT equivalent to the statement ``the particle is in the eigenstate $x_0$''. The second will be true only if the whole function is at $x_0$. That is, if the $x_0$ is the support of the wave-function. The first is a lot less stringent: the wave-function can even be distributed over an infinite range (e.g. a Gaussian wave packet) just as long the average is $x_0$. This means that we can have statements like ``the expected position $\langle x \rangle$ of the particle is exactly $x_0$''$\wedge$ ``the expected momentum $\langle p \rangle$ of the particle is exactly $p_0$'' without having contradictions (e.g. a Gaussian wave packet can be centered on any value of position and momentum).

Note that all statements that characterize the position and momentum of the center of mass form a sub-algebra of the Borel algebra. If the potential for a quantum particle can be considered constant over the wave-function (i.e. $\langle V(x) \rangle = V(\langle x \rangle)$ ), the motion of the center of mass will reduce to the case of a point-particle. In fact, this is exactly how the point particle approximation arises in classical mechanics as well: when we study the motion of planets, cannonballs or specs of dust, we understand that these objects are not point-like, but they have an extent and a matter distribution. The point-particle approximation works insofar we are interested in the motion of the center of mass and the forces are such that the extent of the object does not matter. Naturally, if the orbit of a planet would be such that its distance is less than the radius of the sun, the point-particle approximation would fail spectacularly. In this regard, quantum mechanics works the same way.

Now that we have shown that quantum mechanics can be given an algebra that looks more like the one of classical mechanics, let us do the converse: show that we can (at least formally) give classical mechanics an algebra that looks like the one of quantum logic. As we said before, classical objects are not point-like. So let us consider the space of all possible matter distributions $\rho(x,p)$ over phase-space. We can formalize this as a Hilbert space, where a vector $\psi$ corresponds to $\sqrt{\rho(x,p)}$. This will be a real vector space, and we are only going to consider operators that can be expressed as a multiplicative function $f(x,p)$. Note that $\langle \psi | F \psi \rangle = \int f(x,p) \rho(x,p) dx dp$ as one would expect.\footnote{To be clear: we are not arguing that this construction is the most appropriate or useful. Simply that it can be done.}  We can now take the algebra of subspaces and, as in the quantum case, it will be non-distributive: the disjunction of two subspaces will not be the set union, but the span.

To simplify, let us assume that there are only two possible states $1$ and $2$. A state is identified by a two-valued distribution $[\rho_1, \rho_2]$, which we can take it to be either the probability of the particle being in each state or the amount of material in each state. We note $[k,0]$ the subspace spanned by the first state only. This corresponds to the case where the whole system is in state $1$. Conversely, $[0,k]$ corresponds to the case where the whole system is in state $2$. The disjunction $[k,0] \vee [0,k]$ will correspond to the state spanned by both, which is the whole space $[k,j]$. We can also take the subspace $[k,k]$ which is the subspace spanned by vectors that have equal components in $1$ and $2$: the system is equally distributed in the two cases. Note that $[k,0]$, $[0,k]$ and $[k,k]$ are all pairwise disjoint: no pair have a vector in common apart from zero. Therefore their conjunctions $[k,0] \wedge [0,k] = [0,k] \wedge [k,k] = [k,k] \wedge [k,0] = [0,0]$ are the zero dimensional. On the other hand, $[k,k]$ is a subspace of the whole space $[k,j]$ and therefore $[k,k] \wedge [k,j] = [k,k]$. Putting it all together, we have $[k,k] \wedge ( [k,0] \vee [0,k] ) = [k,k] \wedge [k,j] = [k,k]$. On the other hand, $( [k,k] \wedge [k,0] ) \vee ( [k,k] \wedge [0,k] ) = [0,0] \vee [0,0] = [0,0]$. Therefore the lattice is not distributive.

The lack of distributivity has nothing to do with the complexity of the Hilbert space or non-commuting observables. It's a property of the lattice of subspaces of a vector spaces. In fact, it's a property of the lattice on subgroups more in general. In the subgroup lattice, the disjunction is the span, so doing the disjuction first makes our set bigger during the calculation. Naturally, one may still argue that it is appropriate to focus on this lattice in quantum mechanics while it is not appropriate for classical mechanics. But one cannot argue that this difference is a fundamental difference in the mathematical or logical structure of the theories. Certainly, one cannot say that quantum mechanics requires a departure from classical mechanics when both theories follow similar structures.

The key difference between the algebras of statements, then, is not really about the type of the logical structure: it is its content in terms of statements. If we take two intervals $U_x$ and $U_p$ in position and momentum, and form the statement ``the state is wholly in $U_x$ and $U_p$'', meaning that the vector components are zero outside of those regions, in the classical case we always find at least one $\rho(x,p)$ that satisfies the bounds, while in quantum mechanics, if the bounds violate the uncertainty principle, no suitable $\psi(x)$ can be found. It is the content of the logic that changes.

>>>>

\section{Math}
\label{Math}

In this section we present our main argument, which is that the state space of quantum mechanics and the space of classical probability distributions possess a parallel structure that allows one to apply commutative propositional calculus (i.e. ``classical logic'') on the first and non-commutative propositional calculus (i.e. ``quantum logic'') on the second. That is, both theories will have two lattices of propositions: a commutative one that follows classical logic and a non-commutative one that follows quantum logic.

Note that the classical space we will be using is not phase-space itself, but the space of probability distributions over phase-space, which is a real Hilbert space and thus has a similar structure to the quantum counterpart. The central point is that many features of quantum logic stem not from quantum mechanics per-se, but from the fact that quantum mechanics is a probabilistic theory, which means a classical probabilistic theory will share the same features. Critically, we will see that the non-commutativity of the quantum logic lattice does not originate from non-commuting observables or superposition of quantum mechanics: it is a property of the lattice of subspaces of any vector space and, indeed, of the lattice of sugroups of any group more in general.

Mathematically, the classical commutative lattice will be provided by the Borel algebra, the smallest $\sigma$-algebra that contains the topology; the non-classical non-commutative lattice will be provided by the set of subgroups ordered by inclusion. We will see that the non-commutative lattice is a strict subset of the classical commutative one, meaning that everything that can be described in quantum logic can also be described in classical logic. Moreover, the commutative lattice will include propositions, like propositions on the expectation value of observables, that are of scientific interest.\footnote{In an experimental context, in fact, by ``measurement'' one typically refers to the statistical collection of many ``single takes''. Quantum logic describes the single-take measurements. For example, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves.}

While topologies and $\sigma$-algebras are foundational structures in mathematics for differential geometry,[CITE: Lee, Introduction to Smooth Manifolds] measure theory and probability theory,[CITE: Grimmet/Stirzaker, Probability and Random Processes] they do not seem well-known among physicists and philosophers of science, therefore we briefly summarize their definition, use and meaning. [CITE: Kelly, Logic of Reliable Inquiry - Carcassi/Aidala, Assumptions of Physics]

We start with a set $X$ which can be thought to represent the set of all possible cases we need to distinguish: all quantum states in the quantum case; all possible probability distributions in the classical case; for a simpler example, the set of the real numbers for a continuous quantity. Both topologies and $\sigma$-algebras are collections of subsets of $X$. For example, the standard topology for the real numbers will contain all the open intervals $(a, b)$. The difference between a topology and a $\sigma$-algebra is what types of set operations one can take within them.

A collection of subsets may seem quite an abstract concept, but it can be readily made concrete: any statement on the objects represented by $X$ can be represented by a set. For example, the open set $(a, b)$ will exactly correspond to the statement ``the value of the quantity is between $a$ and $b$''. It is therefore useful to think about both the topology and the $\sigma$-algebra as collections of statements about $X$ instead of sets.

A $\sigma$-algebra is closed under countable union, countable intersection and complement with respect to $X$, meaning that performing those operations on the sets contained in a $\sigma$-algebra will return another set in the same $\sigma$-algebra. In terms of statements, this means that it is closed under countable conjunction, countable disjunction and complement: the $\sigma$-algebra can be thought as a Boolean algebra of statements. Note that this interpretation is exactly the one used in probability. In fact, a probability space is defined by three things: a sample space $\Omega$ which represents the set of all possible outcomes; a $\sigma$-algebra over $\Omega$ which represents all events, all propositions; and a measure that assigns a probability to each event.

A topology, instead, is closed under arbitrary union and finite intersection only. In terms of the statements, it is closed under arbitrary disjunction and finite conjunction. A topology is therefore not a Boolean algebra, but a Heyting algebra. Heyting algebras play a central role in intuitionist logic (and in computer science) as they describe lattices of provable (or computable) statements in finite time. In physics, we can understand the topology to represent statements that are experimentally verifiable in finite time.[CITE again Kelly and Carcassi/Aidala] For example, the open interval $(a, b)$ is an open set precisely because the statement ``the value of the quantity is between $a$ and $b$'' is something we can verify experimentally. Note that the sets in a topology are called open sets precisely because they are a generalization of the open intervals. On the other hand, a set with only one real value $\{ x_0 \}$ is not an open set, it is not in the topology, because the statement ``the value of the quantity is exactly $x_0$'' is not verifiable experimentally due to infinite precision.

Establishing a topology, then, means establishing what statements are assumed to be experimentally verifiable. Given a topology, its Borel algebra is the smallest $\sigma$-algebra that contains all sets in the topology. That is, it is the smallest Boolean algebra that contains all the verifiable statements associated to the topology. Topologies and $\sigma$-algebra are pervasive in physics, to the point that all physically interesting spaces are, directly or indirectly, topological spaces. And wherever we have a topology, we have a Borel algebra, which is a $\sigma$-algebra, a Boolean algebra.

On complete normed spaces, such as Hilbert spaces, there is a standard way to construct the topology. Let us see first how it works in a standard three-dimensional Euclidean space. Given two points $x$ and $y$, we can write the square of distance $\epsilon^2 = |y - x|^2$ as the square of the norm of the vector between the two points. The statement ``the position is within $\epsilon$ meters of $x$'' corresponds to the set $U = \{y \; | \; |y - x|^2 < \epsilon^2 \}$ of all the points whose distance from $x$ is less than $\epsilon$. These are exactly the open sets that generate the topology. This works in the same way for all normed spaces. A Hilbert space, in particular, comes with an inner product which also defines a norm. We can define a distance $\epsilon^2 = |\psi - \phi|^2=\langle \psi - \phi , \psi - \phi \rangle$ between two states, and the statement ``the state of the system is within $\epsilon$ of $\psi$'' corresponds to the set $U = \{\phi \; | \; |\psi - \phi|^2 < \epsilon^2\}$. As we'll see later, the space of classical probability distributions is also a Hilbert space where each point corresponds to the square root of the density. We will have a distance $\epsilon^2 = |\sqrt{\rho_1} - \sqrt{\rho_2}|^2=\langle \sqrt{\rho_1} - \sqrt{\rho_2} , \sqrt{\rho_1} - \sqrt{\rho_2} \rangle$ and we can proceed in the same exact way. In all normed spaces, the topology is constructed from sets of the form ``the object is within $\epsilon$ of $x$''.\footnote{In the case of Hilbert spaces and quantum mechanics, the map between topology and experimental verification does not work perfectly because absolute modulus and phase are not physically meaningful. This can be fixed by using the corresponding projective space, which removes the overdetermination. We are not going to do this as projective spaces have their own peculiarities and are used less.}

From this topology, one can construct the Borel algebra. The Borel algebra also contains the closed sets, which are defined to be the complement of open sets. Recall that, in quantum logic, the propositions corresponded to closed subspaces, which are closed sets and are therefore Borel sets. All the propositions of quantum logic are therefore readily found in the Borel algebra of the Hilbert space. The Borel algebra, which is a Boolean algebra that follows classical logic, contains also all the statements of quantum logic. This construction also applies identically for the classical cases, where we will also be able to talk about the set of all closed subspaces which will have a parallel structure.

While all of this may seem daunting to someone who has never seen this material, we want to stress that these are very standard constructions. We are not taking some obscure framework from some fringe field of mathematics: these are the standard mathematical techniques used to study these spaces and can be found in any textbook. What we are doing here is bringing out their full physical significance, which is too often overlooked. Understanding the link between the mathematical structures and the physics they represent is crucial to clearing out possible misunderstandings.

Now that we have briefly seen the definitions and basic constructions, we can gain more insights into these structures and what they represent. First, let us understand why the lattice of subspaces (i.e. quantum logic) fails to be a Boolean algebra. Stone's representation theorem tells us that any Boolean algebra can be expressed as an algebra of sets using the standard set operations (i.e. intersection, union and complement). For the lattice of subspaces, note that the disjunction $p \vee q$ does not correspond to the set union of the subspaces, but the span. In fact, the union of subspaces is not part of the lattice.  This is why the lattice, even if it is a complemented $0-1$ lattice, fails to be a Boolean algebra: it does not use the standard set operations.

Clearly, $\sigma$-algebras are Boolean algebras because they are exactly an algebra of sets with standard set operations. As we said before, the Borel algebra will contain the lattice of subspaces plus all the set operations between them, including the union. In a way, the lattice of subspaces is not Boolean because it does not contain enough sets, which are instead contained in the Borel algebra.

One may ask: do these extra sets correspond to physically meaningful propositions? To answer that, let us first demonstrate the following proposition.

\begin{prop}
	Let $H$ be a Hilbert space and $\langle \, , \, \rangle$ be its inner product. Let $X$ be a self-adjoint bounded linear operator. Let $F_X : H \to \mathbb{R}$ be the map defined by $F_X(\psi) = \langle \psi , X \psi \rangle$. Let $U \subseteq \mathbb{R}$ be a Borel set. Then $F_X^{-1}(U)$ is a Borel set.
\end{prop}

Proof. Let $X$ be a self-adjoint bounded linear operator. Recall that a linear operator is continuous if and only if it is bounded. Therefore $X$ is continuous. Note that $F_X$ is the composition of $X$ with the inner product. Recall that the inner product is a continuous function. Since $F_X$ is the composition of continuous functions it is also continuous. Recall that all continuous functions are also Borel measurable, therefore $F_X$ is Borel measurable. Let $U \subseteq \mathbb{R}$ be a Borel set. Since $F_X$ is Borel measurable, the reverse image $F_X^{-1}(U)$ is also a Borel set. This concludes the proof.

@@

Borel sets also allow us to express statements such as ``the expected position $\langle x \rangle$ of the particle is exactly $x_0$''. This is NOT equivalent to the statement ``the particle is in the eigenstate $x_0$''. The second will be true only if the whole function is at $x_0$. That is, if the $x_0$ is the support of the wave-function. The first is a lot less stringent: the wave-function can even be distributed over an infinite range (e.g. a Gaussian wave packet) just as long the average is $x_0$. This means that we can have statements like ``the expected position $\langle x \rangle$ of the particle is exactly $x_0$''$\wedge$ ``the expected momentum $\langle p \rangle$ of the particle is exactly $p_0$'' without having contradictions (e.g. a Gaussian wave packet can be centered on any value of position and momentum).

We can go one step further. Note that all statements that characterize the position and momentum of the center of mass form a sub-algebra of the Borel algebra. Recall that the Ehrenfest theorem states that $m \frac{d}{dt}\langle x \rangle = \langle p \rangle$ and $ \frac{d}{dt}\langle p \rangle = - \langle V(x) \rangle$. If the potential for a quantum particle can be considered constant over the wave-function (i.e. $\langle V(x) \rangle = V(\langle x \rangle)$ ), the motion of the center of mass will reduce to the case of a point-particle. The Borel algebra contains as a sub-algebra the statements we need for (at least one form of) the classical limit. And this is a purely classical subalgebra, because the average position and average momentum can both be well defined. Moreover, note that this is not dissimilar to what happens in classical mechanics when, for example, we study planetary motion using only the position and momentum of the center of mass. The algebra of statements for the position and momentum of the center of mass can be independently described and studied. In this regard, the logical structure of the Borel algebra of quantum mechanics works in the same way.

Now that we have shown that quantum mechanics can be given an algebra that looks more like the one of classical mechanics, let us do the converse: show that we can (at least formally) give classical mechanics an algebra that looks like the one of quantum logic. Given that quantum mechanics is a probabilistic theory, let us compare it to a classical probabilistic theory. So let us consider the space of all possible probability distributions $\rho(x,p)$ over phase-space. We can formalize this as a Hilbert space, where a vector $\psi$ corresponds to $\sqrt{\rho(x,p)}$. This will be a real vector space, and we are only going to consider operators that can be expressed as a multiplicative function $f(x,p)$. Note that $\langle \psi | F \psi \rangle = \int f(x,p) \rho(x,p) dx dp$ as one would expect.\footnote{To be clear, we are not arguing that this construction is the most appropriate or useful. Simply that it can be done.}  We can now take the algebra of subspaces and, as in the quantum case, it will be non-distributive: the disjunction of two subspaces will not be the set union, but the span.

To really understand the source of the non-commutativity, let us study the simplest possible example. Let us assume that there are only two possible classical states, $1$ and $2$. A state is identified by a two-valued distribution $[\rho_1, \rho_2]$, where $\rho_1$ is the probability to find the object in state 1 and $\rho_2$ is the probability to find the object in state 2. We note $[k,0]$ the subspace spanned by the first state only. This corresponds to the statement ``the system is certainly in state $1$'' since the probability for state $2$ is zero. Conversely, $[0,k]$ corresponds to the statement ``the system is certainly in state $2$''. The disjunction $[k,0] \vee [0,k]$ will correspond to the subspace spanned by both, which is the whole space $[k,j]$ where we can assign different non-zero probability to either case. We can also take the subspace $[k,k]$, the subspace spanned by vectors that have equal components in $1$ and $2$, which corresponds to the statement ``the system is equally likely to be in either state $1$ or $2$''. Note that $[k,0]$, $[0,k]$ and $[k,k]$ are all pairwise disjoint: no pairs have a vector in common apart from zero. Therefore their conjunctions $[k,0] \wedge [0,k] = [0,k] \wedge [k,k] = [k,k] \wedge [k,0] = [0,0]$ are zero dimensional. On the other hand, $[k,k]$ is a subspace of the whole space $[k,j]$ and therefore $[k,k] \wedge [k,j] = [k,k]$. Putting it all together, we have $[k,k] \wedge ( [k,0] \vee [0,k] ) = [k,k] \wedge [k,j] = [k,k]$. But we also have $( [k,k] \wedge [k,0] ) \vee ( [k,k] \wedge [0,k] ) = [0,0] \vee [0,0] = [0,0]$. Therefore the lattice is not distributive.

Note that during the discussion there was no mention as to whether $k$ and $j$ where real numbers, complex numbers or quaternions. The argument is formally the same in all cases. Therefore the lack of distributivity has nothing to do with the complexity of the Hilbert space or non-commuting observables. It's a property of the lattice of subspaces of any vector space and, more generally, of the lattice of subgroups of any group. In fact, consider the group of boost transformations in relativity. The boosts along one direction form a subgroup. If we take the subgroups of boost along $x$ and $y$ respectively, the smallest subgroup that contain both, the disjunction, is the subgroup of all boosts along any direction in the $x$-$y$ plane. As before, the disjunction is the span, so the resulting set is different from the union, and the union of all boosts along the x-axis and the y-axis is not a subgroup and therefore not contained in the lattice. Thus, this has the same non-commutative structure as quantum logic.

We want to be clear that we are not claiming that the lattice of subspaces of quantum logic is not of interest, or that it is of equal interest in both classical and quantum mechanics. We are pointing it out that the argument that quantum mechanics REQUIRES a radical departure from classical logic in untenable, given both theories follow similar structures. Such arguments have to be a lot more nuanced. Additionally, once we understand the similarity between the theories, we are in a much better position to understand the differences, which naturally are there.

For example, quantum mechanics can only be expressed as a probabilistic theory and therefore ALWAYS requires a structure that in classical mechanics is not mandatory. In quantum mechanics, all subspaces of dimension one represent a pure state while in classical mechanics this is not the case. If we take two intervals $U_x$ and $U_p$ in position and momentum, and form the statement ``the state is wholly in $U_x$ and $U_p$'', meaning that the vector components are zero outside of those regions, in the classical case we always find at least one $\rho(x,p)$ that satisfies the bounds, while in quantum mechanics, if the bounds violate the uncertainty principle, no suitable $\psi(x)$ can be found. The point is that the differences are in the content of the logical structure, on the content of the statements, not on the type of structure.
